PyPy's Approach to Virtual Machine Construction

### 2006-06-01 talk/dls2006/paper.tex
Final tweaks.

Our tool-chain is capable of translating the Python interpreter of
PyPy, written in RPython, currently producing either ANSI C code as
described before, or LLVM\footnote{The LLVM project is the realization
of a portable assembler infrastructure, offering both a virtual
machine with JIT capabilities and static compilation. Currently we are
using the latter with its good high-level optimizations for PyPy.}
assembler which is then compiled to native code with LLVM tools.

The tool-chain has been tested with and can sucessfully apply
transformations enabling various combinations of features. The
translated interpreters are benchmarked using pystone (a Dhrystone 2.0
\cite{dhry20} derivative traditionally used by the Python community,
although it is a rather poor benchmark) and the classical Richards
benchmark (ported to Python) and compared against CPython 2.4.3 \cite{cpy243}
results and are summarized in table \ref{perfsumm}.

@@

{\bf pypy-c:}
    The simplest variant; translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC.

{\bf pypy-c-thread:}
    The same, with OS thread support enabled (thread support is kept
    separate for measurement purposes because it has an impact on the
    GC performance).

{\bf pypy-c-stackless:}
    The same as pypy-c, plus the ``stackless transformation'' step which
    modifies the flow graph of all functions to allow them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework:}
    In this variant, the ``gc transformation'' step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework:}
    This variant combines the ``gc transformation'' step with the
    ``stackless transformation'' step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

{\bf pypy-llvm-c:}
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

{\bf pypy-llvm-c-prof:}
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30\% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is manages to be a bit worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.

@@

As described in section \ref{experimentalresults}, the performance of
the compiled Python interpreters is still not up to competing with the
well-established CPython.  We are always working to improve matters,
considering new optimizations and better GCs.  Also, the OOTyper and
back-ends for CLI/.NET and Smalltalk/Squeak are currently in progress.

@@

To achieve high performance for dynamic languages such as Python, the
proven approach is to use dynamic compilation techniques, i.e.\ to write
JITs.  With direct techniques, this is however a major endeavour, and
increases the effort involved in further evolution of the language.

@@

Targetting platforms that are very different from C/Posix is work in
progress, but given that many of the initial components are shared with
the existing stack of transformations leading to C, we are confident
that this work will soon give results.  Moreover, we believe that these
results will show reasonable efficiency, because the back-ends for VMs
like Squeak and .NET can take advantage of high-level input (as opposed
to trying to translate, say, C-like code to Smalltalk).

>>>>

Our tool-chain is capable of translating the Python interpreter of
PyPy, written in RPython, currently producing either ANSI C code as
described before, or LLVM\footnote{
The LLVM \cite{LLVM:CGO04} project is the realization
of a portable assembler infrastructure, offering both a virtual
machine with JIT capabilities and static compilation. Currently we are
using the latter with its good high-level optimizations for PyPy.}
assembler which is then compiled to native code with LLVM tools.

The tool-chain has been tested with and can sucessfully apply
transformations enabling various combinations of features. The
translated interpreters are benchmarked using pystone (a Dhrystone 2.0
\cite{dhry20} derivative traditionally used by the Python community,
although it is a rather poor benchmark) and the classical Richards
benchmark (ported to Python) and compared against CPython 2.4.3 \cite{cpy243}.
Results are summarized in table \ref{perfsumm}.

@@

{\bf pypy-c:}
    the simplest variant; translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC.

{\bf pypy-c-thread:}
    the same, with OS thread support enabled (thread support is kept
    separate for measurement purposes because it has an impact on the
    GC performance).

{\bf pypy-c-stackless:}
    the same as pypy-c, plus the ``stackless transformation'' step which
    modifies the flow graph of all functions to allow them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework:}
    in this variant, the ``gc transformation'' step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework:}
    this variant combines the ``gc transformation'' step with the
    ``stackless transformation'' step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

{\bf pypy-llvm-c:}
    the same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

{\bf pypy-llvm-c-prof:}
    the same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than a more customized approach;
kernel-level profiling shows that pypy-c typically spends 30\% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC manages to be a bit worse.  The interaction with processor caches is
also hard to predict and account for; in general, we tend to produce
quite large amounts of code and prebuilt data.  The overall performance
is still reasonable: some variants are within the same order of
magnitude as CPython, while others trade a slow-down for functionalities
not available in CPython.

@@

As described in section \ref{experimentalresults}, the performance of
the compiled Python interpreters is not yet up to competing with the
well-established CPython.  We are always working to improve matters,
considering new optimizations and better GCs.

Current work also includes the OOTyper and back-ends for CLI/.NET and
Smalltalk/Squeak, all in-progress.

@@

To achieve high performance for dynamic languages such as Python, the
proven approach is to use dynamic compilation techniques, i.e.\ to write
JITs.  With direct techniques, this is a major endeavour, and
increases the effort involved in further evolution of the language.

@@

Targetting platforms that are very different from C/Posix is work in
progress, but given that many of the initial components are shared with
the existing stack of transformations leading to C, we are confident
that this work will soon give results.  Moreover, we believe that these
results will show reasonable efficiency, because our back-ends for VMs
like Squeak and .NET can take advantage of a high-level input (as opposed
to trying to translate, say, C-like code to Smalltalk).
